---
title: "Data preprocessing"
author: "Irene García- Espantaleón"
date: "2025-06-19"
output: html_document
---

This .Rmd contains the preprocessing of the dataset that will be annotated for the
analysis. I drew a sample of around 15,000 texts from the original 41,000, and 
aggregated the target variable. At the end of the document, I export the processed
dataset that will be annotated by the LLMs.

## Setup and libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(readr)
library(here)
library(purrr)
```

## Preprocessing

Load dataset, available at [this
repo](https://github.com/bvidgen/Dynamically-Generated-Hate-Speech-Dataset/tree/main).

```{r}
data <- read.csv(here("data", "hatespeech_dataset.csv"))
# 41,144 rows

glimpse(data)
```

Basic info on the data from the repo:

-   'acl.id' is the unique ID of the entry.

-   'Text' is the content which has been entered. All content is
    synthetic.

-   'Label' is a binary variable, indicating whether or not the content
    has been identified as hateful. It takes two values: hate, nothate.

-   'Type' is a categorical variable, providing a secondary label for
    hateful content. For hate it can take five values: Animosity,
    Derogation, Dehumanization, Threatening and Support for Hateful
    Entities. Please see the paper for more detail. For nothate the
    'type' is 'none'. In round 1 the 'type' was not given and is marked
    as 'notgiven'.

-   'Target' is a categorical variable, providing the group that is
    attacked by the hate. It can include intersectional characteristics
    and multiple groups can be identified. For nothate the type is
    'none'. Note that in round 1 the 'target' was not given and is
    marked as 'notgiven'.

-   'Level' reports whether the entry is original content or a
    perturbation.

-   'Round' is a categorical variable. It gives the round of data entry
    (1, 2, 3 or 4) with a letter for whether the entry is original
    content ('a') or a perturbation ('b'). Perturbations were not made
    for round 1.

-   'Round.base' is a categorical variable. It gives the round of data
    entry, indicated with just a number (1, 2, 3 or 4).

-   'Split' is a categorical variable. it gives the data split that the
    entry has been assigned to. This can take the values 'train', 'dev'
    and 'test'. The choice of splits is explained in the paper.

-   'Annotator' is a categorical variable. It gives the annotator who
    entered the content. Annotator IDs are random alphanumeric strings.
    There are 20 annotators in the dataset.

-   'acl.id.matched' is the ID of the matched entry, connecting the
    original (given in 'acl.id') and the perturbed version.

Shorthand labels for the target identities:

-   none -\> for non hateful entries NoTargetRecorded -\> for hateful
    entries with no target recorded

-   mixed.race -\> Mixed race background

-   ethnic.minority -\> Ethnic Minorities

-   indig -\> Indigenous people

-   indig.wom -\> Indigenous Women

-   non.white -\> Non-whites (attacked as 'non-whites', rather than
    specific non-white groups which are generally addressed separately)

-   trav -\> Travellers (including Roma, gypsies)

-   bla -\> Black people

-   bla.wom -\> Black women

-   bla.man -\> Black men

-   african -\> African (all 'African' attacks will also be an attack
    against Black people)

-   jew -\> Jewish people

-   mus -\> Muslims

-   muswom -\> Muslim women

-   wom -\> Women

-   trans -\> Trans people

-   gendermin -\> Gender minorities

-   bis -\> Bisexual

-   gay -\> Gay people (both men and women)

-   gay.man -\> Gay men

-   gay.wom -\> Lesbians

-   dis -\> People with disabilities

-   working -\> Working class people

-   old.people -\> Elderly people

-   asi -\> Asians

-   asi.man -\> Asian men

-   asi.wom -\> Asian women

-   asi.east -\> East Asians

-   asi.south -\> South Asians (e.g. Indians)

-   asi.chin -\> Chinese people

-   asi.pak -\> Pakistanis

-   arab -\> Arabs, including people from the Middle East

-   immig -\> Immigrants

-   asylum -\> Asylum seekers

-   ref -\> Refguees

-   for -\> Foreigners

-   eastern.europe -\> Eastern Europeans

-   russian -\> Russian people

-   pol -\> Polish people

-   hispanic -\> Hispanic people, including latinx and Mexicans

-   nazi -\> Nazis ('Support' type of hate)

-   hitler -\> Hitler ('Support' type of hate)

Selecting the columns I will need:

```{r}
data <- data %>% 
  select(X, text, label, type, target, level, round.base) %>% # keeping meaningful cols
  filter(round.base != 1) %>% # first round has no data on hate target
  select(-round.base) |> 
  filter(level == "original") |> # keeping only original texts
  select(-level) |> 
  filter(type != "support") |> # support has 207 observations
  filter(target != "old.people") # around 20 observations and it's difficult to aggregate
# 14,954 rows
```

Our sample to annotate has 15,000 rows.

Aggregate hate targets:

```{r}
data$target <- str_replace_all(data$target, fixed("notargetrecorded"), "unknown")

# category assignment rules:
assign_categories <- function(labels) {
  # Trim whitespace
  labels <- str_trim(labels)

  # Categories initialization
  gender_labels <- c("wom", "wom", "wom") # "wom" and anything ending with ".wom"
  ethnicity_labels <- c(
    "mixed.race", "ethnic.minority", "indig", "indig.wom", "non.white",
    "bla", "bla.wom", "bla.man", "african", "asi", "asi.wom", "asi.man", "arab", "hispanic", "trav"
  )
  nationality_labels <- c("asi.east", "asi.south", "asi.chin", "asi.pak", "eastern.europe", "russian", "pol", "other.national")
  migrant_labels <- c("immig", "asylum", "ref", "for")
  lgbt_labels <- c("trans", "gendermin", "bis", "gay", "gay.man", "gay.wom", "lgbtq")
  religion_labels <- c("jew", "mus", "mus.wom")
  disability_labels <- c("dis")
  class_labels <- c("wc")

  categories_found <- list(
    gender = FALSE,
    ethnicity = FALSE,
    nationality = FALSE,
    migrant = FALSE,
    lgbt = FALSE,
    religion = FALSE,
    disability = FALSE,
    class = FALSE
  )
  
  unmapped <- character()

  for (lab in labels) {
    lab <- str_trim(lab)
    # Check gender: "wom" or ends with ".wom"
    if (lab == "wom" || str_ends(lab, ".wom")) {
      categories_found$gender <- TRUE
    }
    # Ethnicity exact matches
    if (lab %in% ethnicity_labels) {
      categories_found$ethnicity <- TRUE
    }
    # Nationality exact matches
    if (lab %in% nationality_labels) {
      categories_found$nationality <- TRUE
    }
    # Migrant exact matches
    if (lab %in% migrant_labels) {
      categories_found$migrant <- TRUE
    }
    # LGBT exact matches
    if (lab %in% lgbt_labels) {
      categories_found$lgbt <- TRUE
    }
    # Religion exact matches
    if (lab %in% religion_labels) {
      categories_found$religion <- TRUE
    }
    # Disability exact matches
    if (lab %in% disability_labels) {
      categories_found$disability <- TRUE
    }
    # Class exact matches
    if (lab %in% class_labels) {
      categories_found$class <- TRUE
    }
    # If label didn't match any category, add to unmapped
    if (!(
      lab == "wom" || str_ends(lab, ".wom") ||
      lab %in% ethnicity_labels ||
      lab %in% nationality_labels ||
      lab %in% migrant_labels ||
      lab %in% lgbt_labels ||
      lab %in% religion_labels ||
      lab %in% disability_labels ||
      lab %in% class_labels
    )) {
      unmapped <- c(unmapped, lab)
    }
  }
  
  return(list(categories = categories_found, unmapped = unmapped))
}

# Main function to process dataframe
process_targets <- function(df) {
  # Split labels, map to categories, create binary columns
  df2 <- df %>%
    mutate(
      target_list = str_split(target, ",\\s*")
    ) %>%
    rowwise() %>%
    mutate(
      cat_map = list(assign_categories(target_list))
    ) %>%
    mutate(
      gender = cat_map$categories$gender,
      ethnicity = cat_map$categories$ethnicity,
      nationality = cat_map$categories$nationality,
      migrant = cat_map$categories$migrant,
      lgbt = cat_map$categories$lgbt,
      religion = cat_map$categories$religion,
      disability = cat_map$categories$disability,
      class = cat_map$categories$class,
      unmapped = list(cat_map$unmapped)
    ) %>%
    ungroup()

  # Extract all unmapped labels for review
  unmapped_all <- unique(unlist(df2$unmapped))

  list(processed_data = df2 %>% select(-cat_map, -target_list, -unmapped),
       unmapped_labels = unmapped_all)
}

# Use and test the function
result <- process_targets(data)
result$unmapped_labels

data <- result$processed_data
```

Aggregate nationality and migrant into "foreign":

```{r}
# create a column for none
data <- data |> 
  mutate(none = str_detect(target, fixed("none")))

# drop unknown target rows (35)
data <- data |> 
  filter(target != "unknown")

# aggregate nationality and migrant (all nationalities listed are foreign with
# respect to the authors of the texts

data <- data %>%
  mutate(foreign = nationality | migrant) |>
  select(-nationality, -migrant)
```

Create variable for target_counts:

```{r}
target_counts <- data %>%
  summarise(
    none = sum(none, na.rm = TRUE),
    gender = sum(gender, na.rm = TRUE),
    ethnicity = sum(ethnicity, na.rm = TRUE),
    foreign = sum(foreign, na.rm = TRUE),
    # nationality = sum(nationality, na.rm = TRUE),
    # migrant = sum(migrant, na.rm = TRUE),
    lgbt = sum(lgbt, na.rm = TRUE),
    religion = sum(religion, na.rm = TRUE),
    disability = sum(disability, na.rm = TRUE),
    class = sum(class, na.rm = TRUE)
  )

print(target_counts)
```

Convert type to factor and reorder levels:

```{r}
data$type <- fct_relevel(as.factor(data$type), "none", "animosity", "dehumanization", "derogation", "threatening")
```

Export to .csv format:

```{r}
write.csv(data, here("data", "data.csv"), row.names = FALSE) # already included in the repo
```
